\chapter{Wrap-up}
\label{chap:conclusions}
\section{Conclusion} 
In this thesis, the problem to provide a platform to simulate aggregate systems over a LoRaWAN networks has been addressed.
Firstly, the focus has been on improving and extending the simulator DingNet, refining its model and adding new features.
Then, the support to simulate aggregate systems inside it has been designed. 
During this phase the focus has been on producing a platform that allows to move the simulated system to a real deployment with a low effort.
Finally, two different case studies have been presented to verify all the improvements and extensions introduced.
At the end it is possible to say that the achieved platform enables the simulation of a aggregate system over a LoRaWAN network, and allows to move it in a real deploy in a simple way. 
To do that are necessary only two activities:
\begin{enumerate}
    \item change the time concept used from the Protelis nodes, from the simulator time to the time of the real device that hosts the node
    \item change the MQTT client implementation (if it was used a mock one) and the MQTT broker address to subscribe to receive the mote packets.
\end{enumerate}
 So the entire aggregate system developed for the simulation (Protelis backend and program) does not require any changing.

\section{Future work}
% Starting from the work presented in this thesis are available several improvements and interesting topics in different areas.
Several improvements and interesting topics in different areas are available starting from the work presented in this thesis.

One area is to further improve the DingNet simulator. 
For example, it is possible to refine the model by simulating the loss of synchronism among the devices clock. It is possible to do it with different abstraction levels and required effort; starting with the application of a simple jitter to the clocks, arriving to introduce a stochastic model like a Markov chain.

Another area concerns the case study presented in \Cref{chap:case-staudyAC}. 
Here it is possible to evolve the program applying the techniques of machine learning to predict periods with high pollution and take countermeasures in advance. 
For example, it is possible to train a neural network that predicts the pollution level of the following days, considering the actual situation and the weather forecast.

Finally, an interesting topic deals with an extension of the work presented here.
This thesis is focused on joining the aggregate computing with the communication technology LoRaWAN, and provide a platform to simulate this kind of systems.
But real complex pervasive scenarios do not use usually only one communication technology.
So, to support aggregate applications in real use case scenarios, a middleware is needed to fill the network abstraction gap introduced by the paradigm.
Consequently, the future work is to find an existing middleware that can fill this abstraction gap complying with the aggregate computing requirements. For example, Sentilo could be an interesting starting solution.

\iffalse
Finally, it exists an interesting topic not directly correlated with the DingNet simulator, but more aggregate system.
In the introduction of this thesis has been cited the platform Sentilo, as enabling platform to design pervasive system, abstracting from the communication technology of its devices.
So the topic is to study if it is possible to use the platform Sentilo, or a similar one, to fill part of the abstraction gap introduced from the aggregate computing paradigm.
\fi